Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond

### 2024-04-13 main.tex
:pencil: Update geodesic RCG

		The main idea is to perform \emph{additive} updates along the conjugate direction guided by the Riemannian gradient, then project the solution back onto the manifold.

@@

			\item \emph{Translate to the Riemannian gradient \cite{Absil2009}:} At point $\mathbf{\Theta}^{(r)}$, the Riemannian gradient lies in the tangent space of the Stiefel manifold $\mathcal{T}_{\mathbf{\Theta}_g^{(r)}}\mathbb{U}^{L \times L} \triangleq \{\mathbf{M} \mid \mathbf{M}^\mathsf{H} \mathbf{\Theta}_g^{(r)} + {\mathbf{\Theta}_g^{(r)\mathsf{H}}} \mathbf{M} = \mathbf{0}\}$. It gives the steepest ascent direction of the objective on the manifold can be obtained by projecting the Euclidean gradient onto the tangent space:

@@

			where $\gamma$ is the step size refinable by the Armijo rule \cite{Armijo1966};

@@

@@

		A Lie group is simultaneously a continuous group and a differentiable manifold.
		Lie algebra refers to the tangent space of the Lie group at the identity element.
		The exponential map acts as a bridge between the Lie algebra and Lie group, which allows one to recapture the local group structure using linear algebra techniques.
		% On the other hand, a geodesic is a curve representing the shortest path between two points in a Riemannian manifold.
		% Its tangent vectors remain parallel when transported along the curve and the acceleration vectors stay normal to the motion.
		Since unitary matrices forms a Lie group that is closed under multiplication and inversion, and the corresponding geodesic emanating from $\mathbf{\Theta}_g$ with a velocity matrix $\mathbf{S}$ can be described by \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_g = \exp(t \mathbf{S}) \mathbf{\Theta}_g,
		\end{equation}

		% Lie groups are closed under multiplication and inversion, and the set of unitary matrices $\mathbb{U}^{L \times L}$ forms a Lie group.
		% The set of unitary matrices $\mathbb{U}^{L \times L}$ forms a Lie group that is closed under multiplication and inversion.
		% Since unitary matrices in the Stiefel manifold $\mathbb{U}^{L \times L}$ constitute a Lie group, the corresponding geodesic emanating from $\mathbf{\Theta}_g$ with velocity $\mathbf{S}$ is described by the exponential map \cite{Edelman1998}
		% \begin{equation}
		% 	\mathbf{G}_g = \exp(t \mathbf{S}) \mathbf{\Theta}_g,
		% \end{equation}
		% $\mathbf{\Theta}_g^{(r)}$ with velocity $\mathbf{D}_g^{(r)}$ and step size $\mu$ is described compactly by the exponential map \cite{Edelman1998}

		% A Lie group is a group that is also a differentiable manifold.
		% The Lie group of unitary matrices $\mathbb{U}^{L \times L}$ is a manifold that is closed under multiplication and inversion.
		% A geodesic is a curve representing the shortest path between two points in a Riemannian manifold.
		% locally minimizes the distance between two points in a Riemannian manifold.

		% Inspired by \cite{Abrudan2008,Abrudan2009}, we propose a group-wise \gls{rcg} algorithm along the geodesics on the Lie group of unitary matrices $\mathbb{U}^{L \times L}$.

		% It is worth noticing that the feasible domain

		% It is worth noting that the objective \eqref{ob:pareto} is smooth in $\mathbf{\Theta}$ and the feasible domain \eqref{co:pareto_unitary} corresponds to the Stiefel manifold for group $g$.

		% Both properties are often satisfied in \gls{bd}-\gls{ris} optimization problems, and there exists two general solutions in the literature:
		% \begin{itemize}
		% 	\item \emph{Quasi-Newton method \cite{Shen2020a}:}
		% 	% This is a general-purpose optimization algorithm that approximates the Hessian matrix by the gradient and iteratively updates the scattering matrix along the negative gradient direction.
		% \end{itemize}

		% have exploited either the quasi-Newton method on the reactance
		% Instead of solving it directly, we zoom out to general \gls{bd}-\gls{ris} optimization problem and propose an efficient and universal solution.
		% Next, we zoom out to general smooth maximization problems of asymmetric \gls{bd}-\gls{ris}.

		Inspired by \cite{Abrudan2008,Abrudan2009}, we propose a group-wise \gls{rcg} algorithm along the geodesics on the Lie group of unitary matrices $\mathbb{U}^{L \times L}$.
		It leverages the fact that unitary matrices are closed under multiplication.
		At iteration $r$, the gradient is computed in the Euclidean space and translated to the Riemannian manifold \cite{Absil2009}
		% \begin{gather}
		% 	\nabla_{\mathrm{E},g}^{(r)} = \frac{\partial f(\mathbf{\Theta}_g^{(r)})}{\partial \mathbf{\Theta}_g^*},\label{eq:gradient_euclidean}\\
		% 	\nabla_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{E},g}^{(r)} {\mathbf{\Theta}_g^{(r)\mathsf{H}}} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)\mathsf{H}}}.\label{eq:gradient_riemannian} % $\mathcal{O}(N_\mathrm{S}^3)$
		% \end{gather}
		% The Polak-Ribierre parameter \cite{Polak1969} is approximated as \cite{Abrudan2009}
		% \begin{equation}
		% 	\gamma_g^{(r)} = \frac{\mathrm{tr}\bigl((\nabla_{\mathrm{R},g}^{(r)} - \nabla_{\mathrm{R},g}^{(r-1)}) {\nabla_{\mathrm{R},g}^{(r)\mathsf{H}}}\bigr)}{\mathrm{tr}\bigl(\nabla_{\mathrm{R},g}^{(r-1)} {\nabla_{\mathrm{R},g}^{(r-1)\mathsf{H}}}\bigr)}, % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
		% 	\label{eq:parameter_cg}
		% \end{equation}
		% and the conjugate direction is
		% \begin{equation}
		% 	\mathbf{D}_g^{(r)} = \nabla_{\mathrm{R},g}^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)}. % $\mathcal{O}(N_\mathrm{S}^2)$
		% 	\label{eq:direction_cg}
		% \end{equation}
		on the Stiefel manifold, the geodesic emanating from $\mathbf{\Theta}_g^{(r)}$ with velocity $\mathbf{D}_g^{(r)}$ and step size $\mu$ is described compactly by the exponential map \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_g^{(r)}(\mu) = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)}. % $\mathcal{O}(N_\mathrm{S}^3)$
			\label{eq:geodesic}
		\end{equation}
		An appropriate $\mu^\star$ can be obtained by the Armijo rule \cite{Armijo1966}.\footnote{To double the step size, one only need to square the rotation matrix instead of recomputing the matrix exponential, i.e., $\exp(2 \mu \mathbf{D}_g^{(r)}) = \exp^2(\mu \mathbf{D}_g^{(r)})$.}
		Finally, the scattering matrix is updated along the geodesic as
		\begin{equation}
			\mathbf{\Theta}_g^{(r+1)} = \mathbf{G}_g^{(r)}(\mu^\star).
			\label{eq:update_geodesic}
		\end{equation}

		It is worth mentioning that the updates can be performed in parallel for all groups, which is particularly suitable for large-scale \gls{bd}-\gls{ris} optimization problems.
		One can also avoid group-wise updates by treating the entire scattering matrix as a single optimization variable, but this may lead to higher computational complexity and slower convergence.

		Algorithm~\ref{ag:rcg} summarizes the proposed group-wise geodesic \gls{rcg} method for smooth maximization problems of asymmetric \gls{bd}-\gls{ris}.
		Convergence to stationary points is guaranteed.

		\begin{remark}
			Compared with universal manifold optimization \cite{Absil2009,Pan2022d}, Algorithm~\ref{ag:rcg} inherits a trifold benefit from \cite{Abrudan2008,Abrudan2009}:
			\begin{enumerate}
				\item No retraction thanks to rotational update \eqref{eq:geodesic}, \eqref{eq:update_geodesic};
				\item Lower computational complexity per iteration;
				\item Faster convergence thanks to proper parameter space.
			\end{enumerate}
		\end{remark}

		\setalgorithmcaptionfont{\small}
		\begin{algorithm}[!t]
			\small
			\caption{Group-wise geodesic \gls{rcg} for asymmetric \gls{bd}-\gls{ris}}
			\label{ag:rcg}
			\begin{algorithmic}[1]
				\Require $f(\mathbf{\Theta})$, $G$
				\Ensure $\mathbf{\Theta}^\star$
				\Initialize {$r \gets 0$, $\mathbf{\Theta}^{(0)}$}
				\Repeat
					\For {$g \gets 1$ to $G$}
						\State $\nabla_{\mathrm{E},g}^{(r)} \gets$ \eqref{eq:gradient_euclidean} \label{ln:gradient_euclidean}
						\State $\nabla_{\mathrm{R},g}^{(r)} \gets$ \eqref{eq:gradient_riemannian}
						\State $\gamma_g^{(r)} \gets$ \eqref{eq:parameter_cg}
						\State $\mathbf{D}_g^{(r)} \gets$ \eqref{eq:direction_cg}
						\If {$\Re\bigl\{\mathrm{tr}({\mathbf{D}_g^{(r)\mathsf{H}}} \nabla_{\mathrm{R},g}^{(r)})\bigr\} < 0$} \Comment{not an ascent direction}
							\State $\mathbf{D}_g^{(r)} \gets \nabla_{\mathrm{R},g}^{(r)}$
						\EndIf
						\State $\mu \gets 1$
						\State $\mathbf{G}_g^{(r)}(\mu) \gets$ \eqref{eq:geodesic}
						\While {$f\bigl(\mathbf{G}_g^{(r)}(2\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) \ge \mu \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
							\State $\mu \gets 2 \mu$
						\EndWhile
						\While {$f\bigl(\mathbf{G}_g^{(r)}(\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) < \mu / 2 \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
							\State $\mu \gets \mu / 2$
						\EndWhile
						\State $\mathbf{\Theta}_g^{(r+1)} \gets$ \eqref{eq:update_geodesic}
					\EndFor
					\State $r \gets r+1$
				\Until $\lvert f(\mathbf{\Theta}^{(r)}) - f(\mathbf{\Theta}^{(r-1)}) \rvert / f(\mathbf{\Theta}^{(r-1)}) \le \epsilon$
			\end{algorithmic}
		\end{algorithm}
	\end{subsection}

>>>>

		The main idea is to perform {additive} updates along the conjugate direction guided by the Riemannian gradient, then {project} the solution back onto the manifold.

@@

			\item \emph{Translate to the Riemannian gradient \cite{Absil2009}:} At point $\mathbf{\Theta}^{(r)}$, the Riemannian gradient lies in the tangent space of the Stiefel manifold $\mathcal{T}_{\mathbf{\Theta}_g^{(r)}}\mathbb{U}^{L \times L} \triangleq \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} \mathbf{\Theta}_g^{(r)} + {\mathbf{\Theta}_g^{(r)\mathsf{H}}} \mathbf{M} = \mathbf{0}\}$. It gives the steepest ascent direction of the objective on the manifold can be obtained by projecting the Euclidean gradient onto the tangent space:

@@

			where $\mu$ is the step size refinable by the Armijo rule \cite{Armijo1966};

@@

				\label{eq:add_then_retract}

@@

		A Lie group is simultaneously a continuous group and a differentiable manifold.
		Lie algebra refers to the tangent space of the Lie group at the identity element.
		The exponential map acts as a bridge between the Lie algebra and Lie group, which allows one to recapture the local group structure using linear algebra techniques.
		Note that the set of unitary matrices $\mathbb{U}^{L \times L}$ forms a Lie group $U(L)$ under multiplication, and the corresponding Lie algebra $\mathfrak{u}(L) \triangleq \mathcal{T}_{\mathbf{I}}\mathbb{U}^{L \times L} = \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} + \mathbf{M} = \mathbf{0}\}$ consists of skew-Hermitian matrices.
		A geodesic emanating from the identity with velocity $\mathbf{D} \in \mathfrak{u}(L)$ can be described by \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_\mathbf{I}(\mu) = \exp(\mu \mathbf{D}),
			\label{eq:geodesic_identity}
		\end{equation}
		where $\exp(\mathbf{A}) = \sum_{k=0}^\infty (\mathbf{A}^k/k!)$ is the matrix exponential and $\mu$ is the step size (i.e., magnitude of the tangent vector).
		Note that the right translation is an isometry in $U(L)$.
		During the optimization of group $g$, the geodesic evaluated at the identity \eqref{eq:geodesic_identity} should be translated to $\mathbf{\Theta}_g^{(r)}$ for successive updates \cite{Abrudan2008}
		\begin{equation}
			\mathbf{G}_g^{(r)}(\mu) = \mathbf{G}_\mathbf{I}(\mu) \mathbf{\Theta}_g^{(r)} = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:geodesic_translated}
		\end{equation}
		while the Riemannian gradient evaluated at $\mathbf{\Theta}_g^{(r)}$ \eqref{eq:gradient_riemannian} should be translated back to the identity for exploiting the Lie algebra \cite{Abrudan2008}
		\begin{equation}
			\tilde{\nabla}_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{R},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} = \nabla_{\mathrm{E},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)\mathsf{H}}}.
			\label{eq:gradient_translated}
		\end{equation}
		After gradient translation, the deviation parameter and conjugate direction can be determined similarly to \eqref{eq:parameter_cg} and \eqref{eq:direction_cg}
		\begin{equation}
			\tilde{\gamma}_g^{(r)} = \frac{\mathrm{tr}\bigl((\tilde{\nabla}_{\mathrm{R},g}^{(r)} - \tilde{\nabla}_{\mathrm{R},g}^{(r-1)}) {\tilde{\nabla}_{\mathrm{R},g}^{(r)\mathsf{H}}}\bigr)}{\mathrm{tr}\bigl(\tilde{\nabla}_{\mathrm{R},g}^{(r-1)} {\tilde{\nabla}_{\mathrm{R},g}^{(r-1)\mathsf{H}}}\bigr)}. % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
			\label{eq:parameter_cg_geodesic}
		\end{equation}
		\begin{equation}
			{\mathbf{D}}_g^{(r)} = \tilde{\nabla}_{\mathrm{R},g}^{(r)} + \tilde{\gamma}_g^{(r)} {\mathbf{D}}_g^{(r-1)},
			\label{eq:direction_cg_geodesic}
		\end{equation}
		The solution can thus be updated along the geodesic in a multiplicative rotational manner
		\begin{equation}
			\mathbf{\Theta}_g^{(r+1)} = \mathbf{G}_g^{(r)}(\mu) = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:update_geodesic}
		\end{equation}
		where an appropriate $\mu$ may be obtained by the Armijo rule.
		To double the step size, one can simply square the rotation matrix instead of recomputing the matrix exponential, since $\exp^2(\mu \mathbf{D}_g^{(r)}) = \exp(2 \mu \mathbf{D}_g^{(r)})$.

		\setalgorithmcaptionfont{\small}
		\begin{algorithm}[!t]
			\small
			\caption{Group-wise geodesic \gls{rcg} for \gls{bd}-\gls{ris} design}
			\label{ag:rcg}
			\begin{algorithmic}[1]
				\Require $f(\mathbf{\Theta})$, $G$
				\Ensure $\mathbf{\Theta}^\star$
				\Initialize {$r \gets 0$, $\mathbf{\Theta}^{(0)}$}
				\Repeat
					\For {$g \gets 1$ to $G$}
						\State $\nabla_{\mathrm{E},g}^{(r)} \gets$ \eqref{eq:gradient_euclidean} \label{ln:gradient_euclidean}
						\State $\tilde{\nabla}_{\mathrm{R},g}^{(r)} \gets$ \eqref{eq:gradient_translated}
						\State $\tilde{\gamma}_g^{(r)} \gets$ \eqref{eq:parameter_cg_geodesic}
						\State $\mathbf{D}_g^{(r)} \gets$ \eqref{eq:direction_cg_geodesic}
						\If {$\Re\bigl\{\mathrm{tr}({\mathbf{D}_g^{(r)\mathsf{H}}} \tilde{\nabla}_{\mathrm{R},g}^{(r)})\bigr\} < 0$} \Comment{not an ascent direction}
							\State $\mathbf{D}_g^{(r)} \gets \nabla_{\mathrm{R},g}^{(r)}$
						\EndIf
						\State $\mu \gets 1$
						\State $\mathbf{G}_g^{(r)}(\mu) \gets$ \eqref{eq:geodesic_translated}
						\While {$f\bigl(\mathbf{G}_g^{(r)}(2\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) \ge \mu \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
							\State $\mu \gets 2 \mu$
						\EndWhile
						\While {$f\bigl(\mathbf{G}_g^{(r)}(\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) < \mu / 2 \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
							\State $\mu \gets \mu / 2$
						\EndWhile
						\State $\mathbf{\Theta}_g^{(r+1)} \gets$ \eqref{eq:update_geodesic}
					\EndFor
					\State $r \gets r+1$
				\Until $\lvert f(\mathbf{\Theta}^{(r)}) - f(\mathbf{\Theta}^{(r-1)}) \rvert / f(\mathbf{\Theta}^{(r-1)}) \le \epsilon$
			\end{algorithmic}
		\end{algorithm}

		Algorithm~\ref{ag:rcg} summarizes the proposed \gls{bd}-\gls{ris} design framework based on group-wise geodesic \gls{rcg}.
		Compared to the non-geodesic approach, our method leverages the properties of unitary group to replace the add-then-retract update \eqref{eq:add_then_retract} with a multiplicative rotational update \eqref{eq:update_geodesic} along the geodesic.
		This leads to faster convergence and simplifies the step size tuning thanks to an appropriate parameter space.

		Both approaches can be performed in parallel for all groups to facilitate large-scale \gls{bd}-\gls{ris} optimization problems.
		Since block-unitary matrices are also closed under multiplication, one can avoid group-wise updates by operating directly on $\mathbf{\Theta}$ and pinching (i.e., keeping the block diagonal and nulling the other entries) the Euclidean gradient \eqref{eq:gradient_euclidean}, at the cost of higher computational complexity and slower convergence.
	\end{subsection}
\end{section}
